Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data

### 2014-03-07 doc/2014bioinformatics_zhang/RVD3.tex
2014-03-05-RVD3_sensitivity_analysis, and updated draft.

\begin{abstract}
%Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
%To address the detection of SNVs, a variety of methods are being under-represented.
%However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.

%We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately. We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior function.
%Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
%Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
%For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
%This model is able to detect the emergence of a beneficial SNV earlier than was previously shown.
%Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data

% The same with the Abstract poster of RECOMB2014.
Massively parallel sequencing data generated by next-generation sequencing (NGS) technology is routinely used to detect single nucleotide variants (SNVs) in research samples.
An emerging challenge for this technology is the identification of SNVs in heterogeneous cell populations with low read-depth data.
We have developed a Bayesian statistical model is able to share information between correlated positions and call low-frequency variants in heterogeneous samples.
We present a Bayesian sensitivity analysis of the model to variations in the prior function.
Our model with different priors both performs a high accuracy, and a Jeffrey's prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with an improper prior.
In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Priors for precision parameter
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Priors for precision parameter}
The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful. A way to choose the prior function is to use the Schwarz criterion or Bayesian information criterion (BIC) \citep{weakliem1999critique}. Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}. To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust}. Based on this theory, we analyze the posterior and prior probability of the precision hyper-parameter.
For this purpose, a bunch of different priors should be chosen for analysis \citep{gelman2006prior}. In a polygenic modeling with Bayesian sparse linear mixed model research, the results don't change dramatically by changing the prior measurement which reveals that the prior specification for the hyper-parameters are fine \citep{zhou2013polygenic}. So we performed three different plausible prior distributions: improper prior, informative prior (log-normal), and non-informative prior (Jeffrey's).

@@

The Bayesian posterior predictive distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
These plots indicate the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
The positions are chosen in the middle of the base length- position 104 and 244 are mutant, and position 100 and 300 are non-mutant.
The posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
Generally the distribution plots display normal and stable without a peak nor a strange shape.
The two distributions show different ways of the prior and posterior.
The log-normal prior attributes a higher prior probability to M values between 0 and 2000.
The Jeffrey's prior assigns more information than the log-normal prior (flat) from the posterior curve.
They both want to search for a small value for M from the prior curve.

The proper prior (log-normal) and non-informative prior (Jeffrey¡¯s) are performed to compare with the improper prior. There is little variability across positions indicating that the replication variance does not change greatly across position. We show several key parameters of the model in Figure~\ref{fig:priors_compare}. A key model parameter M is less variable with a log-normal and improper prior compared to the Jeffrey¡¯s prior.
Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots. $M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.

\begin{figure*}[!tbp]
\begin{center}
\includegraphics[width=185mm]{figs/post_prior_dilu_10.pdf}
\caption{Distribution of priors and posteriors when dilution is 10\%}
\label{fig:dilu_10}
\end{center}
\end{figure*}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=160mm]{figs/priors_compare.pdf}
\caption{Key parameters for RVD3 model with different priors.}
\label{fig:priors_compare}
\end{center}
\end{figure}

@@

Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
Log-normal prior shows a higher sensitivity and specificity value than the Jeffrey's prior.
Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others.
It is obvious no matter Jeffrey's or log-normal prior, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors, which seems desirable of our model for extending for a broad application.
Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
Here we chose Jeffrey's prior model because it's a non-informative prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=80mm]{tables/Sen_Speci.pdf}
\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
\label{fig:SS}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=80mm]{tables/FDR.pdf}
\caption{False Discovery Rate comparison with different priors on RVD3.}
\label{fig:fdr}
\end{center}
\end{figure}

>>>>

\begin{abstract}

% The same with the Abstract poster of RECOMB2014.
Massively parallel sequencing data generated by next-generation sequencing (NGS) technology is routinely used to detect single nucleotide variants (SNVs) in research samples.
An emerging challenge for this technology is the identification of SNVs in heterogeneous cell populations with low read-depth data.
We have developed a Bayesian statistical model is able to share information between correlated positions and call low-frequency variants in heterogeneous samples.
We present a Bayesian sensitivity analysis of the model to variations in the prior function.
Our model with different priors both performs a high accuracy, and a Jeffrey's prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with the log-normal prior and an improper prior. 
And log-normal prior performs a higher specificity than the other two. 
In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Priors for precision parameter
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Priors for precision parameter}
The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful. A way to choose the prior function is to use the Schwarz criterion or Bayesian information criterion (BIC) \citep{weakliem1999critique}. Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}. To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust, harrison2013bayesian}. Based on this theory, we analyze the posterior and prior probability of the precision hyper-parameter.
For this purpose, a bunch of different priors should be chosen for analysis \citep{gelman2006prior}. In a polygenic modeling with Bayesian sparse linear mixed model research, the results don't change dramatically by changing the prior measurement which reveals that the prior specification for the hyper-parameters are fine \citep{zhou2013polygenic}. So we performed three different plausible prior distributions: improper prior, informative prior (log-normal), and non-informative prior (Jeffrey's).

@@

The Bayesian posterior distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
These plots indicate the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
The positions are chosen in the middle of the base length- position 185 and 265 are mutant, and position 200 and 300 are non-mutant.
The posterior probability distributions for $M_j$ are estimated by Gaussian Kernel Density \citep{silverman1986density}.
Generally the distribution curves display normal and stable.
Jeffrey's posterior distribution changes a lot which shows Jeffrey's piror is more sensitive than the log-normal prior between the mutant and non-mutant positions.

Figure~\ref{fig:priors_compare} shows that across all priors the non-mutant positions indicate that the replicate variance is small. 
At the known mutant positions, the replicate parameter $M_j$ is less variable compared to the reference position with the log-normal and improper prior compared with Jeffrey's prior, which corresponds to the information in Figure~\ref{fig:dilu_10}
However, at the begining and the end of the sequence which are the low read-depth positions, the improper and log-normal prior are more variable compared to Jeffrey's prior.
Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots. 
$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision between positions. 
Here $\widetilde{E}[M|D]$ represents the expectation of empirical posterior approximation.

\begin{figure*}[!tbp]
\begin{center}
\includegraphics[width=170mm]{figs/post_prior_10.pdf}
\caption{Distribution of priors and posteriors when MAF is 10\%}
\label{fig:dilu_10}
\end{center}
\end{figure*}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=140mm]{figs/priors_compare.pdf}
\caption{Precision parameter for RVD3 model with different priors. $M_0$ is the global precision. X indicates mutant position.}
\label{fig:priors_compare}
\end{center}
\end{figure}

@@

Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors.
Log-normal prior performs a higher sensitivity and specificity value, and the specificity of log-normal prior is a little better than the improper prior and much better than the performance of Jeffrey's prior.
It shows that log-normal prior performs a smaller false positive rate than the other two priors.

Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others over all the MAF levels at various read depths.
It is obvious no matter Jeffrey's or log-normal prior, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors, which seems desirable of our model for extending for a broad application.
Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter to accommodate to the target. When we do the variants calling research, we should choose Jeffrey's prior model because it's a non-informative prior, and can produce a smaller false discovery rate and high accuracy; when we do the clinical experiment or diagnosis, we prefer to choose log-normal prior which cares more on true positive rate and true negative rate.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=110mm]{tables/Sen_Speci.pdf}
\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
\label{fig:SS}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=100mm]{tables/FDR.pdf}
\caption{False Discovery Rate comparison with different priors on RVD3.}
\label{fig:fdr}
\end{center}
\end{figure}
